Some examples in extended logic counting classes

### 2016-12-29 LICS/preliminaries.tex

\subsection{Function complexity classes}
Assume that $\R = \{R_1, \ldots, R_k\}$ is a relational signature and $\A$ is an ordered finite $\R$-structure with a domain $A$ containing $n$ elements. Recall that  $<$ is a linear order on $A$, say $a_1 < a_2 < \ldots < a_n$. For every $i \in \{1, \ldots, k\}$, define the encoding of $R_i^\A$, denoted by $\enc(R_i^\A)$, as the following binary string. Assume that $\ell = \arity(R_i)$ and consider an enumeration of the $\ell$-tuples over $A$ in the lexicographic order induced by $<$ (that is, $(a_1, \ldots, a_1, a_1)$, $(a_1, \ldots, a_1, a_2)$, $\ldots$, $(a_n, \ldots, a_n, a_{n-1})$, $(a_n, \ldots, a_n, a_n)$). Then let $\enc(R_i^\A)$ be a binary string of length $n^\ell$ such that the $i$-th bit of $\enc(R_i^\A)$ is 1 if the $i$-th tuple in the previous enumeration belongs to $R_i^\A$, and 0 otherwise. Moreover, define the encoding of $\A$, denoted by $\enc(\A)$, as the following binary string~\cite{L04}:
\begin{eqnarray*}
	\enc(\A) & = & 0^n \, 1 \, \enc(R_1^\A) \, \cdots \, \enc(R_k^\A).
\end{eqnarray*}
Given a relational signature $\R$,  function $f$ is said to be {\em defined over $\R$} (or just {\em over $\R$}) if $f$ is a total function from $\{\enc(\A) \mid \A \in \ostr[\R]\}$ to $\nat$. Moreover, a set $\CC$ of functions is said to be a {\em function complexity class} if every $f \in \CC$ is a function defined over some relational signature $\R$.

>>>>

\subsection{Function complexity classes}

### 2016-12-29 LICS/saluja.tex

In the next result, we show that $\eqso$ also defined a finite hierarchy similar than in~\cite{SalujaST95} that we called the $\eqso$-hierarchy.
\begin{proposition}
\[
\QE{0} \; \subsetneq \; \QE{1} \; \subsetneq \; \QU{1} \; \subsetneq \; \QE{2} \; \subsetneq \; \QU{2} \; = \; \eqso(\fo)
\]
\end{proposition}
A natural question until this point is what is the connection between $\E{i}$ and $\U{i}$ and their corresponding classes $\QE{i}$ and $\QU{i}$. Clearly, we have that $\E{i} \subseteq \QE{i}$ and $\U{i} \subseteq \QU{i}$. Indeed, each formula $\varphi(\bar{X}, \bar{x})$ in $\E{i}$ or $\U{i}$ is equivalent to $\sa{\bar X} \sa{\bar x} \varphi(\bar{X}, \bar{x})$ which is in $\QE{i}$ or $\QU{i}$, respectively. 
Then it is left to know whether the containment between the sharp and quantitative classes is strict or not. 
For this, we study first whether each formula $\alpha$ in $\eqso(\LL)$ can be transformed in a \emph{prefix normal form} where all sums quantifiers are restricted to the beginning of the formula. 
Formally, we say that a formula $\alpha$ in $\eqso(\LL)$ is in \emph{$\Sigma$-prenex normal form} (or just prenex normal form) if $\alpha$ is of the form:
\[
\alpha := \sa{\bar{X}} \sa{\bar{x}} \varphi(\bar{X}, \bar{x})
\]
where $\bar{X}$ and $\bar{x}$ is a sequence of $0$ or more second and first order variables, respectively.
In particular, note that a boolean formula is always in $\Sigma$-prenex normal form.
Then we say that a formula $\alpha$ is in \emph{sum normal form} if $\alpha$ is of the form $c + \Sigma_{i=1}^n \alpha_i$ where $c$ is a non-negative constant and each $\alpha_i$ is in $\Sigma$-prenex normal form.
Next we show that each formula in $\eqso(\LL)$ can be converted in \emph{sum normal form} but not always in prenex normal form.
\begin{theorem}
	Every formula in $\eqso(\LL)$ can be rewritten in \emph{sum normal form}. Furthermore, there exists a formula $\alpha \in \loge{1}$ such that there does not exist a formula in $\Sigma$-prenex normal form in $\QE{1}$ equivalent to $\alpha$.
\end{theorem}

>>>>

In the next result, we show that $\eqso$ also defined a finite hierarchy similar than in~\cite{SalujaST95} that we called the $\eqso$-hierarchy.
\begin{proposition}
\[
\QE{0} \; \subsetneq \; \QE{1} \; \subsetneq \; \QU{1} \; \subsetneq \; \QE{2} \; \subsetneq \; \QU{2} \; = \; \eqso(\fo)
\]
\end{proposition}
A natural question until this point is what is the connection between $\E{i}$ and $\U{i}$ and their corresponding classes $\QE{i}$ and $\QU{i}$. Clearly, we have that $\E{i} \subseteq \QE{i}$ and $\U{i} \subseteq \QU{i}$. Indeed, each formula $\varphi(\bar{X}, \bar{x})$ in $\E{i}$ or $\U{i}$ is equivalent to $\sa{\bar X} \sa{\bar x} \varphi(\bar{X}, \bar{x})$ which is in $\QE{i}$ or $\QU{i}$, respectively. 
Then it is left to know whether the containment between the sharp and quantitative classes is strict or not. 
For this, we study first whether each formula $\alpha$ in $\eqso(\LL)$ can be transformed in a \emph{prefix normal form} where all sums quantifiers are restricted to the beginning of the formula. 
Formally, we say that a formula $\alpha$ in $\eqso(\LL)$ is in \emph{$\Sigma$-prenex normal form} (or just prenex normal form) if $\alpha$ is of the form:
\[
\alpha := \sa{\bar{X}} \sa{\bar{x}} \varphi(\bar{X}, \bar{x})
\]
where $\bar{X}$ and $\bar{x}$ is a sequence of $0$ or more second and first order variables, respectively.
In particular, note that a boolean formula is always in $\Sigma$-prenex normal form.
Then we say that a formula $\alpha$ is in \emph{sum normal form} if $\alpha$ is of the form $c + \Sigma_{i=1}^n \alpha_i$ where $c$ is a non-negative constant and each $\alpha_i$ is in $\Sigma$-prenex normal form.
Next we show that each formula in $\eqso(\LL)$ can be converted in \emph{sum normal form} but not always in prenex normal form.
\begin{theorem}
	Every formula in $\eqso(\LL)$ can be rewritten in \emph{sum normal form}. Furthermore, there exists a formula $\alpha \in \QE{1}$ such that there does not exist a formula in $\Sigma$-prenex normal form in $\QE{1}$ equivalent to $\alpha$.
\end{theorem}
